\subsection{\mwp-Bounds Analysis}\label{subsec:mwp}

The \mwp-flow analysis~\cite{jones2009} is a static analysis technique for evaluating resource usage of imperative programs.
It analyzes input variables' value growth, and aims to discover a polynomially bounded data-flow relation between variables \emph{initial} values and \emph{final} values. 
When all variables are bounded by polynomials in inputs, the analysis succeeds, \ie a program is derivable in the underlying calculus.
If one or more variables is not bounded, such bound cannot be established.
The soundness theorem of the analysis guarantees that if a derivation exists, the program's value growth is polynomially bounded in inputs. Furthermore, as a sound technique, it offers a computational method to certify programs data size growth properties at runtime.

\input{figures/pbounds}

Instead of manual inspection, the \mwp-flow analysis provides an \emph{automatable} technique to distinguish programs that have polynomial growth bounds. 
Internally is works my applying inference rules to program commands and tracking variable dependencies by coefficients (or \emph{flows}).
A detailed description of the system is provided in the original work by Jones and Kristiansen~\cite{jones2009} and refined by Aubert et. al~\cite{aubert2022b}. The remainder of this section elaborates on the central notions and terminology to understand the mechanics of the analysis, as it relates to the publications in \autoref{sec:published-manuscripts}.

