%! suppress = LineBreak
%! suppress = TooLargeSection
\section{Seeking Strong Software Quality Guarantees}
\label{sec:aicc-intro}

The gold standard in formal verification is showing that an implementation of a
program is {totally correct}\index{total correctness}. Total correctness
requires proving two conditions about the program: that the program meets its
specification, and that the computation terminates~\cite{leino2023}. The former
ensures the program computes the expected output for all inputs. The latter
ensures that the computed result will become observable {eventually}. Thus, a
proof of total correctness gives us strong assurance that the program behaves
expectedly at runtime, \ie when the program is executed.

An additional consideration in reasoning about program correctness concerns
{non-functional} properties\index{non-functional properties}. Non-functional
properties characterize the program's quality features like execution latency,
power consumption, and information flow security~\cite{terbeek2018}. Pure
functional correctness is \emph{practically unsatisfactory} if a program
requires exorbitant time to compute its result; aggressively drains a device
battery, or reveals secret information~\cite{heraud2011,aubert20222}.
Engineering truly high-quality programs thus requires additional techniques that
enable us to formally verify non-functional properties. Unfortunately, analyzing
non-functional properties is non-trivial and generally
undecidable~\cite{rice1953}\index{undecidability}. Developing specification
formalisms and reasoning methods, that can provide some guarantees despite the
undecidability, is an active area of research~\cite{etaps2025}.

The field of \emph{Implicit Computational Complexity (ICC)}~\cite{dallago2011}%
\IfFileExists{icc}{---\cf\input{icc.tex}\autoref{fig:icc_systems}---}{ }offers a
conceivable pathway toward formal verification of certain non-functional
properties. ICC is complementary to the classic complexity theory because it
uses programming languages as the {mechanism} to reason about runtime behaviors.
Broadly, complexity theory is concerned with resource requirements of
computations~\cite{goldreich2008}. Those requirements are organized in a
hierarchy of {complexity classes}. The classes describe resource requirements
\wrt an abstract (mathematical) machine model. The classes correspond roughly to
resources like running time and memory consumption of programs that are executed
on real-world computers. To streamline this reasoning, \emph{implicit}
computational complexity aims to discover \emph{machine-independent}
characterizations of the complexity classes. ICC systems are designed by
introducing a \emph{restriction}, at the level of a programming language, that
guarantees a complexity property. Then, every program satisfying the restriction
is known to belong to a particular complexity class. This in turn provides a
guarantee about the program's runtime resource consumption.

Exchanging machine models for syntactic criteria is powerful in several ways.
For example, it facilitates formal verification of programs. ICC enables
defining {syntactical} specifications of resource consumption, and developing
fully automatable analyses to verify whether a program satisfies such a
specification~\cite{heraud2011}. ICC systems have been designed for numerous
combinations of programming languages (imperative,  lambda-calculus, function
algebras, quantum, term rewrite systems, \(\pi\)-calculus, \ldots),
restriction-techniques (linear logic, data-flow analysis, type systems, category
theory, \ldots), and complexity classes (\textsc{p}, \textsc{pspace},
\textsc{l}, \textsc{pp}, \ldots)~\cite{moyen2017,pchoux2020}.

%! suppress = TooLargeSection
\section{Addressed Problem and Goals}
\label{sec:aicc-goals}

Besides uses in formal verification, there are many good motivations for
implicit reasoning about computational complexity. ICC allows guaranteeing
program properties \emph{by construction}, before any program exists. ICC drives
better understanding of complexity classes and yields more natural definitions
and proofs of central results than classical complexity
theory~\cite{kristiansen2017}. For example, ICC systems show that complexity
classes are intrinsic mathematical entities that do not depend on a particular
machine model. Multiple ICC systems, \eg~\cite{jones2009,marion2011}, are rich
enough to express natural algorithms. Since the reasoning techniques differ from
the existing alternatives, the results ICC systems produce are often
orthogonal~\cite{aubert20222}. Finally, many ICC
systems---\eg~\cite{jones2009,marion2011,hainry2023,atkey2024}---are
\emph{compositional}. This means they use modular techniques that enable
bottom-up reasoning about subprograms. Compositionality is critical for
scalability~\cite{carbonneaux2015} and useful for proving
soundness~\cite{keidel2021}, \ie that the analysis technique produces correct
results. However, compositionality is missed by many prominent program analysis
techniques~\cite{carbonneaux2015,schiebel2024}.

Despite this rich inventory of compelling features, ICC has remained largely a
theoretical novelty, with a few exceptions%
~\cite{avanzini2017,avanzini2008,moyen20172,hainry2021,hoffmann2012,feree2018}.
A theory in absence of applications means the power, limitations, and utilities
of the theory are difficult to recognize. In our view~\cite[p.
xxxv]{bishop2003}~\cite[p. 75]{moyen2017}, theory and application are
{symbiotic}. Investigations of both are needed for scientific advancement.

The embedded manifesto of the dissertation research is that \emph{applications
are necessary} to push forward our collective understanding of implicit
computational complexity. It is conceivable that potentially great power is
locked inside the theoretical ICC systems. It remains an open problem to release
that power. Within the ICC community, more investigations are needed to explore
the capabilities of the existing systems. Externally, the ICC-based program
analyses should be exposed to broader research communities. These steps would
motivate enhancements in ICC systems and facilitate discovery of new
applications.

The work in the dissertation is only a small step toward these ambitious
long-term goals. The realistic expectations of the dissertation are about
technical and social advancement. Overall, there are four high-level goals.
\begin{enumerate}
\item Extend {applied} capabilities of ICC in automatic program analysis and
verification.
\item Take ICC a few steps closer to integration into real-world software
development workflows.
\item Initiate conversations on the relevance of ICC applications.
\item Expose ICC to broader research communities.
\end{enumerate}
A guiding intuition behind these goals is that, {if applied}, implicit
computational complexity could provide us several new practical program
analyses. The use cases of such analysis include formal verification of many
non-functional properties.

%! suppress = TooLargeSection
\section{Research Questions and Approaches}
\label{sec:aicc-approaches}

The following \emph{main hypothesis} guides the dissertation investigation.
\begin{quote}
\noindent Implicit computational complexity offers applied utilities when lifted
from the theoretical domain.
\end{quote}
Since the hypothesis is broad, the research projects that challenge it must be
more refined. We focus on two paths. The first applies ICC in automatic analysis
of \emph{resource consumption}. This application is natural because it parallels
the theoretical use of ICC. The second path takes the investigation to a more
unconventional direction. It asks whether we can adjust ICC systems to track
\emph{other} program properties, beyond resource consumption. Stated
differently, by swapping the property of interest, it allows us to assess the
technical flexibility of ICC systems. To summarize, we investigate whether we
can \begin{enumerate*}[label=(\roman*)]
\item apply the systems in the originally designed ways, and
\item discover unexpected applications.
\end{enumerate*}

\subsection{Automatic Resource Analysis}
\label{subsec:aicc-automatic-resource-analysis}

This series of work focuses on an ICC system called the \emph{flow calculus of
mwp-bounds}~\cite{jones2009}. The flow calculus is a canonical example of a
theoretical ICC system. It focuses on tracking change in data size during
computation. For imperative programs, it aims to determine if it is possible to
bound the growth of variable values by polynomials in inputs. The choice to
focus on this particular ICC system is based on intuition. As a syntactical and
compositional data-flow analysis, it {seems like} a good candidate for automatic
resource analysis.

Several research questions arise from the flow calculus.

\begin{enumerate}[label={(RQ\arabic*)},leftmargin=*,labelindent=1em]
\item Can we develop an \emph{automatic} program analysis based on the theory?
\item Given its paper proofs, is the theory correct, \ie can we prove formally
the soundness of the flow calculus?
\item Assuming the theory can be automated, what are the use cases of such
analysis?
\end{enumerate}

These questions are still conventional because they target resource analysis. In
other words, given a complexity-theoretic technique, we should reasonably expect
that it can be applied in the designed way. If successful, the investigation
should minimally produce two research artifacts---a program analyzers and
mechanized proof---corresponding to RQ1 and RQ2. Additionally, it should produce
new insights of the flow calculus and its possible applications. Since ICC is
not the only technique for analyzing growth in data size, having an automatic
analyzer would enable comparing it to the alternative approaches.

\subsection{Analyzing Extended Properties}
\label{subsec:extended-props}

The second series of work applies implicit computational complexity to track
{other} non-functional program properties. The idea is to select an ICC system
and adjust it to capture a different non-functional property. The motivation is
multifaceted. It requires developing deep understanding of the mechanics of the
chosen technique. Conceptually, it pushes to think about ICC outside the usual
frame of computational complexity. Lastly, investigations in this direction are
important to expose ICC to broader research communities, since the application
is shifted away from (or beyond) resource consumption.

In this series of work, the starting point is an ICC system based on
quasi-invariants (QI), as formulated in~\cite{moyen20172}. Similar to the flow
calculus, the QI framework is a data-flow analysis of imperative programs.
However, the QI framework supports a richer input language and is founded on a
seemingly adjustable mathematical framework. In the prior
formulation~\cite{moyen20172}, it was used to obtain a compile-time program
optimization. The optimization lifts nested loops outside their containing loop,
such that post-transformation the program complexity is reduced.

We ask the following questions about the extended uses of the QI framework.
\begin{enumerate}[label={(RQ\arabic*)},leftmargin=*,labelindent=1em]
\setcounter{enumi}{3}
\item How to develop a program transformation to increase parallelization potential?
\item How to use it to analyze security properties, specifically non-interference?
\end{enumerate}
Research questions 4 and 5 are posed here with an air of obviousness. However,
arriving to those questions was not obvious. A substantial hidden challenge of
the dissertation research plan involves identifying suitable research
domains---like parallel programming or language-based
security~\cite{sabelfeld2003}---where ICC-based techniques could offer
meaningful benefit.

Another consideration is that adjusting an ICC system may lead to losing the
original guarantee about resource consumption. In return, we gain an alternative
guarantee about program behavior. Whether this trade is beneficial depends on
the application. Therefore, besides tracking properties, this investigation is
also about studying the mechanics of ICC systems, and uncovering how exactly
they provide behavioral guarantees.

Finally, this direction of research contains a bonus\hyp{}challenge. Simply
showing that an ICC-based technique can solve a problem in another application
domain is not enough. It is necessary to show that the solution is
\emph{independently} relevant in the target domain, regardless of the origins of
the solution. Based on experience, this criterion is strongly enforced by the
scientific peer review process.

%! suppress = TooLargeSection
\section{Methodology and Research Deliverables}
\label{sec:aicc-methods}

Following the approaches of~\autoref{sec:aicc-approaches}, we conducted a series of
investigations. These resulted in several
publications~\cite{aubert20222,aubert20232,aubert2023b}, workshop
presentations~\cite{aubert20231,aubert202217,splash22}, and ongoing work.
\IfFileExists{papers}{\input{papers.tex}The dependence associations of the
written works are illustrated in~\autoref{fig:papers}.}{}
We can summarize the main contributions as follows.

For RQ1, we enhanced the theory to develop an automatic resource analysis
in~\cite{aubert20222,aubert2023b,rusch2025}. Due to inherent nondeterminism,
developing an efficient and useful practical analysis was surprisingly
difficult, but eventually achievable. The enhanced flow calculus can answer the
same questions as the original formulation, but automatically and in practice.
Moreover, through the enhancements we discovered how to make the calculus more
expressive, allowing it to cover a larger class of programs than the original
theory~\cite{rusch2025}. The flow calculus is implemented as an open source
static analyzer, pymwp%
\footnote{Pronounced \emph{pa\textsc{i} em double-you p\={e}.}}%
\footnote{https://statycc.github.io/pymwp/}{ }
for analyzing a subset of C programs~\cite{aubert2023b}. Based on this work, we
discovered that the analysis can support inference of specification
conditions~\cite{rusch2025} that are needed for formal verification. This
finding corresponds to RQ3. Proving the soundness of the flow calculus (RQ2) is
still ongoing work~\cite{aubert20232}. However, given the other developments, a
formalization effort is now even more strongly motivated and justified.

Along the second direction, for RQ4, we applied the QI framework to develop a
program optimization for loop parallelization~\cite{aubert20232}. This is useful,
because the optimization can be applied to loops whose iteration count
is unknown. Parallelizing such loops is outside the capabilities of prior
techniques. In an ongoing investigation, for RQ5, we apply the QI
framework to tracking programs {non\hyp{}interference}~\cite{goguen1982}, a kind
of confidentiality property. Non-interference ensures a program does not reveal
secrets; and if it does, we can identify where the violation occurs. Our working
intuition is that, since the analysis can be adjusted to different programming
languages, it enables tracking \emph{preservation} of the non-interference
property at different compilation stages, while a program is transformed from
source code to an executable. This is different from many existing techniques
that analyze programs only at one stage of representation.

%! suppress = TooLargeSection
\section{Research Findings and Conclusions}
\label{sec:aicc-discussion}

\IfFileExists{icc}{\subsubsection*}{\paragraph*}{On automatic resource analysis.}
We can now confirm that it is possible to obtain automatic program analysis from
the flow calculus of mwp-bounds, but only after substantial adjustments. An
efficient program analysis required inventing ways to manage costly
sub-computations and solving additional problems that were exposed in the
process. At conclusion, the enhanced technique is not just automatable, but can
analyze more programs than the original theory. We can now confirm the analysis
results are complementary to alternative state-of-the-art techniques~\cite[p.
5]{aubert2023b}. Thus, it enables us to analyze different questions about
resource consumption. The implemented flow calculus can also be used in
specification inference. There are potentially many more applications of the
flow calculus remaining to be discovered.

\IfFileExists{icc}{\subsubsection*}{\paragraph*}
{On analysis of extended program properties.} The QI framework provides a case
example showing that ICC systems can be flexible enough to track other program
properties. While adjusting the system, we made a surprising discovery in that
changing the property of interest simplified the mathematical analysis. For
example, to track a complexity property, we must compute fixed points and
preserve the program command order. In the adjusted analysis, it was possible to
relax these conditions. More generally, that the QI framework can be used to
track various properties is reminiscent of the Dependency Core Calculus
(DCC)~\cite{abadi1999b}. In DCC, different program properties are modeled as
\emph{instances} of a central notion of \emph{dependency}. This mirrors our
observations about the QI framework. This suggests that the QI framework
encapsulates even wider utility than uncovered so far during this dissertation
work.

\IfFileExists{icc}{\subsubsection*}{\paragraph*}
{On broader dissertation goals.} Zooming out of the specific projects, we can
reflect on the main hypothesis and the research goals. We have successfully
gathered evidence of the application potential of ICC. The evidence supports the
main hypothesis, but is still too limited to draw generalizing conclusions.
However, our experiences of developing the ICC systems strongly reinforced the
symbiotic view of theory and application. In the process, we faced many
scientific and engineering challenges, and identified compelling motivations to
justify the investigations. At conclusion, we have strengthened the theories of
interest and clarified their practical relevance.

In~\autoref{sec:aicc-goals}, we crafted four goals to define the dissertation
expectations. The first was about extending the applied capabilities of ICC,
which was met successfully. The second was about introducing ICC in development
workflows. Although we have developed a static analyzer, it is a step in
\enquote{isolation.} Integrating ICC analyses into other software development
tools (compilers, verifiers, \etc) remains as an outstanding task. Such
integration is important to promote relevance and continue development of ICC
techniques. Our work in using the flow calculus to infer specification
conditions~\cite{rusch2025} is a first step in that direction. The last two
goals were social and about disseminating ideas across research communities.
Within the ICC community, our efforts materialized in seminar talks and informal
conversations. Although this record is below ideal, the ideas in the published
works will remain discoverable to others in written form, asynchronously and in
perpetuity, like~\cite{moyen2017}. In regard to other communities, we were
successful at exposing the work at verification, programming languages, and
security venues. Thus, the fourth goal was satisfactorily met.

\IfFileExists{icc}{\subsubsection*}{\paragraph*}
{Final thoughts \& future perspectives.} The ICC approach of
guaranteeing program properties demonstrates great applied potential. For
example, if offers a pathway toward formal verification of many non-functional
properties. One research direction, that deserves more attention, is applying
ICC systems to ensure properties by construction. This would provide guarantees
before any program exists and eliminates the need for \emph{aposteriori}
analysis. Although the dissertation research did not extend that far, exploring
this direction is an intriguing future goal. Whether it is attainable depends
crucially on a viewpoint shift. We should regard \emph{implicit computational
complexity} in the broader context it offers---including as a rich toolbox of
techniques for static program analysis---and continue the exploration of its
applied potential.

\IfFileExists{acks}{\input{acks}}{}
